---
title: "A3: Incarceration"
output: html_document
---
```{r setup, include=FALSE}
# load summary values
source("summary.R")

# load charts
source("chart1.R")
source("chart2.R")
source("map.R")
```

### Introduction

Racism in the United States has continued to exist since its creation,
surviving far past the end of slavery, segregation, and countless other forms of
discrimination, all of which promised to bring an end to the issue. In the
modern day, this act of injustice manifests itself quite notably within our
prison system, excused as an essential means of upholding a "just" society. In
order to more closely examine and analyze this issue, this page intends to use
the *Incarceration Trends Dataset* to compare the populations of various races
to the incarcerated populations of the same races in order to identify
inequality in the proportion of incarcerated individuals between these groups.
It will also look at how these variables change over time to identify any
historical trends, as well as how they relate to geographical regions to see if
location plays any significant role in the discriminatory incarceration within
the United States.

### Summary Information

Looking into the most complete recent year in the dataset, `r rct_yr`, it was
found that the total percentage of the study population that was incarcerated
was `r total_pct`%, however this wasn't spread proportionately across the five
studied racial groups: Asian American / Pacific Islander, Black, Latinx, Native
American, and White. Rather, Asian Americans / Pacific Islanders had the lowest
ratio, falling significantly below the expected incarceration rate with only
`r min_pct`% of the population being incarcerated, while Black Americans were
imprisoned at the highest rate, well above their expected proportion sitting at
`r max_pct`%. In other words, Black Americans make up only `r black_pop_pct`% of
the total population, yet `r black_inc_pct`% of the incarcerated population.
`r rct_yr` wasn't even the peak of their imprisonment though, that being
`r rct_yr - max_black_yr` years ago in `r max_black_yr` where `r max_black_inc`
Black Americans sat jailed or imprisoned while only `r white_inc_same_yr` White
Americans were incarcerated in the same year.

### The Dataset

This dataset was originally collected by the Vera Institute, and then broken
into smaller files for ease of use in this assignment by Melanie Walsh of UW’s
iSchool. Vera collected this data by using a combination of federal and state
department reports to average the daily amount of prisoners, disaggregated by
race, across each county for an annual report of jail population. Prison
population data was collected from these reports as the number of prisoners at
the end of the year. These results were compiled using sources such as the NCRP,
DCRP, ASJ, COJ, CAF, UCR, CDC, USDA, and U.S. Census. This data was collected,
according to Vera, “to build and create justice systems that ensure fairness,
promote safety, and strengthen communities.” The complete dataset provided by
Vera contains 153,811 observations and 121 features, while the smaller subsets
used in the assignment have 36 and 28 observations for the prison and jail
datasets respectively. One of the main ethical questions that must be considered
when working with this data is ‘what were the circumstances and reasons behind
the incarceration of the individuals within these racial groups,’ as the data
only shows the population incarcerated as a number, not any of the events,
policies, or context behind each person’s incarceration. This doubles as one
of the dataset’s main limitations: it lacks context surrounding the reason of
incarceration. Without this vital information, it is easy to jump to conclusions
about trends in the data, but lacking knowledge of the full circumstances behind
the data, it is impossible to know the full story from this dataset alone.
Another major limitation of the data is the large amount of unknown values,
especially in earlier years, making it difficult to derive any meaningful
observations about the missing populations during those times.

### Trends Over Time Chart

```{r, echo = FALSE, message = FALSE}
chart1
```

This chart displays the percentage of each racial group studied that was
incarcerated between the years 1990 and 2016. Its purpose is to showcase the
discrepancy between the groups, as well as trends within each group as time
passes. For example, Black Americans have a significantly higher percentage of
their population incarcerated across nearly all of the years studied compared to
other groups, while Asian Americans, Pacific Islanders, and White Americans
consistently have a much lower percentage. The chart also effectively displays
major changes in percentage of the populations incarcerated, such as the jumps
in 1992 and falls in 1995 in for Native and Latino populations, or the spike in
1998 for all populations except Latinos. Observing these trends could help link
historical policies or events to their effects on specific racial groups, and
uncover potentially discriminatory and unjust effects not initially apparent.
Overall, the chart shows a general increase in incarcerated populations around
1992, and a decrease beginning after 1998, with much variability in the years
between for most racial groups.

### Variable Comparison Chart

```{r, echo = FALSE, message = FALSE}
chart2
```

This chart shows how the percentage of Black Americans in the total population
relates to the percentage they make up in the incarcerated population of the
same location. Its purpose is to check if Black Americans are being incarcerated
at a level proportional to their share of the population. If this were the case,
the expected visual would have the dots fall on or around a diagonal line where
the percent of the total population is about the same as the percent of the
incarcerated population, displayed above for reference. From the graph, however,
this is clearly not the case, as the vast majority of the dots fall above the
line indicating that they are being incarcerated at a rate disproportional to
the percentage of the total population that they make up. This further supports
the previous results indicating that the United States prison systems could be
discriminating Black Americans, as they are consistently over represented within
the incarcerated populations.

### Map

```{r, echo = FALSE, message = FALSE}
map
```

This map shows which racial group makes up the largest percentage of the
incarcerated population in each county. The purpose of it is to examine if there
are any geographical trends that might suggest discrimination in certain areas.
The results shown indicate that the majority of the United States counties have
White Americans making up most of their incarcerated populations, however there
are some interesting exceptions towards the bottom of the map. Here we can see
that in the southeast section of America, there is a large amount of counties
with Black Americans making up the majority of their incarcerated population,
and in the southwest there is a large percentage made up by Latinos. While this
could just have to do with populations of those groups in those areas, it could
also be a reminder of the conflict and discrimination against those groups in
the United State's history.