---
title: "A3: Incarceration"
output: html_document
---
```{r setup, include=FALSE}
# load summary values
source("summary.R")

# load charts
```

### Introduction

Racism in the United States has continued to exist since its creation,
surviving far past the end of slavery, segregation, and countless other forms of
discrimination, all of which promised to bring an end to the issue. In the
modern day, this act of injustice manifests itself quite notably within our
prison system, excused as an essential means of upholding a "just" society. In
order to more closely examine and analyze this issue, this page intends to use
the *Incarceration Trends Dataset* to compare the populations of various races
to the incarcerated populations of the same races in order to identify
inequality in the proportion of incarcerated individuals between these groups.
It will also look at how these variables change over time to identify any
historical trends, as well as how they relate to geographical regions to see if
location plays any significant role in the discriminatory incarceration within
the United States.

### Summary Information

Looking into the most complete recent year in the dataset, `r rct_yr`, it was
found that the total percentage of the study population that was incarcerated
was `r total_pct`%, however this wasn't spread proportionately across the five
studied racial groups: Asian American / Pacific Islander, Black, Latinx, Native
American, and White. Rather, Asian Americans / Pacific Islanders had the lowest
ratio, falling significantly below the expected incarceration rate with only
`r min_pct`% of the population being incarcerated, while Black Americans were
imprisoned at the highest rate, well above their expected proportion sitting at
`r max_pct`%. In other words, Black Americans make up only `r black_pop_pct`% of
the total population, yet `r black_inc_pct`% of the incarcerated population.
`r rct_yr` wasn't even the peak of their imprisonment though, that being
`r rct_yr - max_black_yr` years ago in `r max_black_yr` where `r max_black_inc`
Black Americans sat jailed or imprisoned while only `r white_inc_same_yr` White
Americans were incarcerated in the same year.

### The Dataset

Who collected the data?  
How was the data collected or generated?  
Why was the data collected?  
How many observations (rows) are in your data?  
How many features (columns) are in the data?  
What, if any, ethical questions or questions of power do you need to consider when working with this data?  
What are possible limitations or problems with this data?   (at least 200 words)


### Trends Over Time Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The first chart that you will create and include will show the trend over time of your variable/topic. Think carefully about what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Show more than one, but fewer than ~10 trends
  - This may mean showing the same measure for different locations or different racial groups. Think carefully about a meaningful comparison of locations (e.g., the top 10 counties in a state, top 10 states, etc.)
- You must have clear x and y axis labels
- The chart needs a clear title 
- You need a legend for your different line colors and a clear legend title

When we say "clear" or "human readable" titles and labels, that means that you should not just display the variable name.

Here's an example of how to run an R script inside an RMarkdown file:

```{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The second chart that you will create and include will show how two different (continuous) variables are related to one another. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- You must have clear x and y axis labels
- The chart needs a clear title 
- If you choose to add a color encoding (not required), you need a legend for your different color and a clear legend title

### Map

Include a chart. Make sure to describe why you included the chart, and what patterns emerged

The last chart that you will create and include will show how a variable is distributed geographically. Again, think carefully about what such a comparison means and what you want to communicate to your user (you may have to find relevant trends in the dataset first!). Here are some requirements to help guide your design:

- Your map needs a title
- Your color scale needs a legend with a clear label
- Use a map based coordinate system to set the aspect ratio of your map

